Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
binop example

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} \} \cup \{ n | n \in
\mathbb N \}$ given in \Cref{fig:dyckinductive}. This grammar is
LR(1), and the corresponding automaton (which is too complex to
include here) uses the $\&$ to guard transitions with lookahead
information \max{Probably we should at least try to fit it in or put
  it in the appendix.}

\begin{theorem}[\agdalogo]
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to a deterministic automaton with lookahead.
\end{theorem}
\begin{figure}
\begin{lstlisting}
data Exp : L where
  num : (*@$\uparrow$@*)(&[ n : Nat ] 'n' (*@$\lto$@*) Exp)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ('+' (*@$\otimes$@*) Atom *) ')' (*@$\lto$@*) Exp)
\end{lstlisting}
%% data Opening : Nat -> Bool -> L where
%% data Closing : Nat -> Bool -> L where
%% data Multiplying : Nat -> Bool -> L where
%%   done : (*@$\uparrow$@*)(Multiplying 0 true)
%%   times : (*@$\uparrow$@*)(&[ n : Nat ] &[ b : Bool] '*' -o Opening n b -o Multiplying n b)
%%   unmatched : (*@$\uparrow$@*)(&[ n : Nat ] Multiplying (n + 1) false)
%%   unexpected
\caption{Binary operation expressions and a corresponding lookahead automaton}
\label{fig:dyckinductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser

>>>>

Our final example is of a simple grammar of arithmetic expressions
with an associative operation. Here we take the alphabet to be $\{
\stringquote{(}, \stringquote{)}, \stringquote{+} \stringquote{NUM}
\}$. In \Cref{fig:binop-inductive} we define it using two mutually
recursive types, corresponding to the two non-terminals we would use
in a CFG syntax. The syntactic structure encodes that the binary
operation is right associative. In the same figure, we define the
traces of an automaton with one token of lookahead. The automaton has
four different ``states'', each with access to a natural number
``stack''. The ``opening'' state $O$ expects either an open paren, in
which case it increments the stack and stays in the opening state, or
sees a number and proceeds to the $D$ state. The ``done opening''
state $D$ is where lookahead is used: if the next token will be a
right paren, then we proceed to $C$, otherwise we proceed to $M$. Here
$\mathsf{NotStartsWithRP}$ is defined as $\I \oplus (('(' \oplus '+'
\oplus \mathsf{NUM}) \otimes \top)$. In the ``closing'' state $C$ if
we observe a close paren we decrement the count and continue to the
$D$ state. In the ``multiplying'' state $M$, we succeed if the string
ends and the count is $0$, and if we see a plus we continue to the $O$
state. Additionally, since the automaton need also parse all of the
incorrect strings, we add additional failing cases. It is
straightforward to implement a parser for this lookahead automaton,
generalizing the approach for deterministic automata.

\begin{theorem}[\agdalogo]
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to $\mathsf{O}~0~\true$ 
\end{theorem}
\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) DO n b (*@$\lto$@*) O n b)
  done : (*@$\uparrow$@*)(&[ n ] O n false)
  unexpected : (*@$\uparrow$@*)(&[ n ] (')' (*@$\oplus$@*) '+') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) O n false)
data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  lookAheadRP  : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot  : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) M n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrow$@*)(&[ n ] ')' (*@$\lto$@*) C 0 b)
  done : (*@$\uparrow$@*)(&[ n ] C n false)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) '+' (*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) C n false)
data M : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  doneGood : (*@$\uparrow$@*)(M 0 true)
  doneBad : (*@$\uparrow$@*)(&[ n ] M (n + 1) false)
  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O n b (*@$\lto$@*) M n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) M n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser
